---
title: "Mixtape Demand Estimation Exercises"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixtape Demand Estimation Exercises}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rblp.verbose = FALSE)
```

## Introduction

This vignette walks through the exercises from the
[Mixtape Sessions Demand Estimation](https://github.com/Mixtape-Sessions/Demand-Estimation)
course, taught by Jeff Gortmaker and others. The original course materials use
Python and [pyblp](https://pyblp.readthedocs.io/en/stable/); here we reproduce
every exercise in R using the **rblp** package.

The exercises build progressively:

1. **Exercise 1 (Pure Logit):** OLS logit, IV logit, fixed effects, counterfactual
   price cuts, and elasticities.
2. **Exercise 2 (Mixed Logit):** Adding demographic interactions and random
   coefficients to capture consumer heterogeneity.
3. **Exercise 3 (Micro Moments):** Using individual-level data to sharpen
   identification (conceptual overview).

The dataset is a simplified version of the Nevo (2000) ready-to-eat cereal data
covering 24 products across 94 city-quarter markets.

## Data Setup

rblp bundles the Mixtape cereal data. The raw data has columns for `market`,
`product`, `mushy`, `servings_sold`, `city_population`, `price_per_serving`,
and `price_instrument`. We use `prepare_mixtape_data()` to compute market shares
and rename columns into the format expected by `blp_problem()`.

```{r data_setup}
library(rblp)

# Load and prepare data
raw <- load_mixtape_products()
products <- prepare_mixtape_data(raw)

cat(sprintf("Product-market observations: %d\n", nrow(products)))
cat(sprintf("Markets (city-quarter): %d\n", length(unique(products$market_ids))))
cat(sprintf("Products per market: %d\n", length(unique(products$product_ids))))
cat(sprintf("Firms: %s\n", paste(sort(unique(products$firm_ids)), collapse = ", ")))
```

### Understanding the Data

The key columns after preparation:

- `market_ids`: city-quarter identifier (e.g., "C01Q1")
- `product_ids`: product identifier (e.g., "F1B04" = Firm 1, Brand 04)
- `shares`: market share = `servings_sold / (city_population * 90)`, where 90
  represents one serving per day over a quarter
- `prices`: price per serving (dollars)
- `mushy`: indicator for whether the cereal is mushy (0/1)
- `firm_ids`: extracted from the first two characters of `product_ids`
- `demand_instruments0`: excluded price instrument for IV estimation

```{r data_summary}
cat("Price per serving summary:\n")
print(round(summary(products$prices), 4))
cat("\nMarket share summary:\n")
print(round(summary(products$shares), 6))
cat("\nMushy breakdown:\n")
print(table(mushy = products$mushy[!duplicated(products$product_ids)]))
```

### Computing the Outside Good Share

In the logit model, the outside good share in each market is
$s_{0t} = 1 - \sum_j s_{jt}$. The outside good represents consumers who do not
purchase any of the 24 inside products.

```{r outside_share}
market_inside_shares <- tapply(products$shares, products$market_ids, sum)

cat("Inside share (sum of product shares per market):\n")
cat(sprintf("  Mean:  %.4f\n", mean(market_inside_shares)))
cat(sprintf("  Min:   %.4f\n", min(market_inside_shares)))
cat(sprintf("  Max:   %.4f\n", max(market_inside_shares)))

cat("\nOutside share:\n")
cat(sprintf("  Mean:  %.4f\n", mean(1 - market_inside_shares)))
cat(sprintf("  Min:   %.4f\n", min(1 - market_inside_shares)))
cat(sprintf("  Max:   %.4f\n", max(1 - market_inside_shares)))
```

The outside share ranges from about 30% to 82%, indicating substantial variation
in how much of each market is captured by the 24 tracked cereal products.

---

## Exercise 1: Pure Logit

### Q1-Q2: Data Description and Market Share Computation

The Mixtape dataset is a panel of 24 cereal products observed in 94
city-quarter markets. For each product-market, we observe the quantity sold
(`servings_sold`), the city population, the price per serving, and one
excluded instrument for price.

Market shares are computed as:
$$s_{jt} = \frac{\text{servings\_sold}_{jt}}{\text{city\_population}_t \times 90}$$

where 90 is the assumed number of potential servings per person per quarter
(about one per day). This defines the total market size, with the outside good
absorbing whatever fraction of the population does not purchase any of the 24
products.

```{r q1_q2}
# Show first few rows of the key variables
head(products[, c("market_ids", "product_ids", "firm_ids",
                   "shares", "prices", "mushy")], 10)
```

### Q3: OLS Logit with `lm()`

The plain logit model has the Berry (1994) linear form:

$$\log(s_{jt}) - \log(s_{0t}) = \beta_0 + \alpha \, p_{jt} + \beta_1 \, \text{mushy}_{jt} + \xi_{jt}$$

We first estimate this by OLS, which is biased because prices $p_{jt}$ are
correlated with the unobserved product-market quality $\xi_{jt}$ (price
endogeneity).

```{r q3_ols}
# Compute the dependent variable: log(s_j) - log(s_0)
outside_shares <- 1 - tapply(products$shares, products$market_ids, sum)
products$s0 <- as.numeric(outside_shares[as.character(products$market_ids)])
products$y <- log(products$shares) - log(products$s0)

# OLS regression
ols_fit <- lm(y ~ prices + mushy, data = products)

cat("=== OLS Logit Results ===\n\n")
print(round(summary(ols_fit)$coefficients, 3))
```

The OLS estimates are:

- Intercept: approximately $-2.935$
- Price coefficient ($\alpha$): approximately $-7.48$
- Mushy: approximately $0.075$ (not statistically significant)

The price coefficient is biased toward zero. High-quality products command
higher prices (positive correlation between $p_{jt}$ and $\xi_{jt}$), which
partially offsets the true negative effect of price on demand.

### Q4: OLS Logit with rblp

We can replicate the OLS logit using `blp_problem()`. The key is to set
`demand_instruments0 = prices`, which makes the instrument set identical to the
regressors (i.e., no excluded instruments), so the IV estimator reduces to OLS.

```{r q4_rblp_ols}
# For OLS: use prices as own instrument (no excluded instruments)
products_ols <- products
products_ols$demand_instruments0 <- products_ols$prices

f1_ols <- blp_formulation(~ prices + mushy)
ols_problem <- blp_problem(list(f1_ols), products_ols)
ols_results <- ols_problem$solve(method = "1s")

cat("=== rblp OLS Logit Results ===\n\n")
print(ols_results)
```

The rblp results match the `lm()` estimates exactly: this confirms that the GMM
framework with the identity instrument matrix reduces to OLS.

### Q5: Fixed Effects Logit (Market + Product FE)

Adding market and product fixed effects absorbs all time-invariant product
characteristics and all market-level demand shifters. This eliminates $\xi_{jt}$
components that are purely product-specific or purely market-specific, leaving
only the within-product, within-market variation for identification.

With fixed effects and prices as their own instrument, this is a "FE logit"
that does not yet address endogeneity of the remaining price variation.

```{r q5_fe}
# FE logit: absorb market + product, prices as own instrument
products_fe <- products
products_fe$demand_instruments0 <- products_fe$prices

f1_fe <- blp_formulation(~ 0 + prices, absorb = ~ market_ids + product_ids)
fe_problem <- blp_problem(list(f1_fe), products_fe)
fe_results <- fe_problem$solve(method = "1s")

cat("=== FE Logit Results (prices as own IV) ===\n\n")
print(fe_results)
```

The price coefficient with market and product fixed effects is approximately
$-28.6$, much more negative than the OLS estimate of $-7.48$. The fixed effects
absorb the product-level quality that was positively correlated with prices,
revealing a much stronger true price sensitivity.

### Q6: IV + Fixed Effects Logit

Now we use the excluded `price_instrument` as an instrument for prices, combined
with the market and product fixed effects. This addresses endogeneity from any
remaining within-product, within-market correlation between prices and
unobserved quality shocks.

```{r q6_iv_fe}
# IV + FE: use the price_instrument as excluded instrument
products_iv <- products
# demand_instruments0 is already set to price_instrument by prepare_mixtape_data()

f1_iv <- blp_formulation(~ 0 + prices, absorb = ~ market_ids + product_ids)
iv_problem <- blp_problem(list(f1_iv), products_iv)
iv_results <- iv_problem$solve(method = "1s")

cat("=== IV + FE Logit Results ===\n\n")
print(iv_results)
```

The IV + FE price coefficient is approximately $-30.6$ (with a standard error
around 1.0). This is slightly more negative than the FE-only estimate,
indicating that even after absorbing fixed effects, there is some remaining
positive correlation between prices and unobserved quality that the instrument
helps correct.

### Comparison of Logit Specifications

```{r comparison}
# ols_results$beta has: (Intercept), prices, mushy
# fe_results$beta and iv_results$beta have: prices only (intercept absorbed)
ols_price_idx <- which(names(coef(ols_fit)) == "prices")
rblp_ols_price <- ols_results$beta[ols_price_idx]  # index 2 = prices

cat("=== Price Coefficient Comparison ===\n\n")
cat(sprintf("  OLS (lm):        %8.3f\n", coef(ols_fit)["prices"]))
cat(sprintf("  rblp OLS:        %8.3f\n", rblp_ols_price))
cat(sprintf("  FE (no IV):      %8.3f\n", fe_results$beta[1]))
cat(sprintf("  FE + IV:         %8.3f  (SE = %.3f)\n",
            iv_results$beta[1],
            iv_results$summary_table()$se[1]))
```

The progression from $-7.5$ (OLS) to $-28.6$ (FE) to $-30.6$ (IV+FE) shows
how each correction --- absorbing fixed effects and instrumenting for price ---
reveals increasingly strong consumer price sensitivity.

### Q7: Price Cut Counterfactual

We now use the IV + FE logit results to simulate a counterfactual: what happens
to market shares if product F1B04 receives a price cut of $0.04 per serving in
market C01Q1? (F1B04's baseline price is about $0.072, so this is roughly a 55%
price reduction.)

In the pure logit, shares have a closed form given mean utilities $\delta_{jt}$.
A price change for product $j$ shifts its mean utility by
$\Delta\delta_j = \alpha \times \Delta p_j$, and new shares follow from the
logit formula.

```{r q7_counterfactual}
# Pick the first market
first_market <- "C01Q1"
mkt_idx <- which(products_iv$market_ids == first_market)
mkt_data <- products_iv[mkt_idx, ]

# Current shares and delta
s0_mkt <- unique(mkt_data$s0)
current_delta <- log(mkt_data$shares) - log(s0_mkt)
current_shares <- mkt_data$shares

# Price coefficient from IV + FE
alpha <- iv_results$beta[1]

# Price cut of $0.04/serving for F1B04
new_prices <- mkt_data$prices
f1b04_idx <- which(mkt_data$product_ids == "F1B04")
price_cut <- 0.04
new_prices[f1b04_idx] <- new_prices[f1b04_idx] - price_cut

cat(sprintf("F1B04 price: $%.4f -> $%.4f (cut of $%.2f/serving)\n",
            mkt_data$prices[f1b04_idx], new_prices[f1b04_idx], price_cut))

# New delta: delta_new = delta_old + alpha * (p_new - p_old)
new_delta <- current_delta + alpha * (new_prices - mkt_data$prices)

# New logit shares
exp_delta <- exp(new_delta)
new_shares <- exp_delta / (1 + sum(exp_delta))

# Percentage change in shares
pct_change <- (new_shares - current_shares) / current_shares * 100

cat("\n=== Counterfactual: $0.04 Price Cut for F1B04 in C01Q1 ===\n\n")
cf_table <- data.frame(
  product = mkt_data$product_ids,
  old_share = sprintf("%.6f", current_shares),
  new_share = sprintf("%.6f", new_shares),
  pct_change = sprintf("%+.1f%%", pct_change),
  stringsAsFactors = FALSE
)
print(cf_table, row.names = FALSE, right = FALSE)

cat(sprintf("\nF1B04 share change: %+.1f%%\n", pct_change[f1b04_idx]))
cat(sprintf("Mean change for other products: %.2f%%\n",
            mean(pct_change[-f1b04_idx])))
```

F1B04's share increases dramatically (by over 200%), while all other products
lose a few percent of their market share. This large response reflects the
estimated price coefficient of approximately $-30.6$: a $0.04 price cut
translates to a utility gain of $0.04 \times 30.6 \approx 1.22$ units, which
is substantial on the logit scale.

In the pure logit, the substitution pattern is governed by the Independence of
Irrelevant Alternatives (IIA) property: all competing products lose share in
proportion to their current shares, regardless of how similar they are to F1B04.

This is both a strength (simplicity) and a weakness (unrealistic substitution
patterns) of the logit model. The random coefficients model in Exercise 2
relaxes IIA.

**Note:** The pyblp Mixtape solutions report F1B04 share change of +223.6% and
other products at -1.45%. Small numerical differences arise from optimizer
tolerances and GMM weighting matrix details.

### Q8: Own-Price Elasticities

The own-price elasticity in the logit model is:
$$\eta_{jj} = \alpha \, p_{jt} \, (1 - s_{jt})$$

Since $\alpha < 0$, own-price elasticities are negative. We compute them using
the `compute_elasticities()` method.

```{r q8_elasticities}
E <- iv_results$compute_elasticities(first_market)
own_elast <- diag(E)

cat("=== Own-Price Elasticities (Market C01Q1) ===\n\n")

elast_table <- data.frame(
  product = mkt_data$product_ids,
  price = sprintf("%.4f", mkt_data$prices),
  share = sprintf("%.6f", mkt_data$shares),
  elasticity = sprintf("%.3f", own_elast),
  stringsAsFactors = FALSE
)
print(elast_table, row.names = FALSE, right = FALSE)

cat(sprintf("\nOwn-price elasticity range: [%.1f, %.1f]\n",
            min(own_elast), max(own_elast)))
cat(sprintf("Mean own-price elasticity: %.2f\n", mean(own_elast)))
```

Own-price elasticities are negative and range from approximately $-2$ to $-5$
(or roughly $-2.4$ to $-6.3$ depending on the exact estimated alpha). More
expensive products have more elastic demand, which makes economic sense: a 1%
price increase for an expensive cereal represents a larger absolute price
change, inducing more consumers to switch away.

### Cross-Price Elasticities

In the logit model, the cross-price elasticity is:
$$\eta_{jk} = -\alpha \, p_{kt} \, s_{kt} \quad (j \neq k)$$

This means all products have the same cross-elasticity with respect to a given
product $k$ --- a direct consequence of IIA.

```{r cross_elasticities}
# Show a 5x5 submatrix of cross-elasticities
cat("=== Cross-Price Elasticity Submatrix (first 5 products) ===\n\n")
sub_E <- E[1:5, 1:5]
rownames(sub_E) <- mkt_data$product_ids[1:5]
colnames(sub_E) <- mkt_data$product_ids[1:5]
print(round(sub_E, 3))

cat("\nNote: Off-diagonal entries in each column are identical (IIA property).\n")
```

---

## Exercise 2: Mixed Logit (Random Coefficients)

The pure logit imposes IIA: substitution patterns are determined entirely by
market shares, not by product similarity. The mixed (random coefficients) logit
relaxes IIA by allowing consumer preferences to vary. Products that are similar
in the characteristic space will naturally be closer substitutes.

### Cross-Market Variation and Identification

A key insight from BLP (1995) is that cross-market variation in product sets
and demographics helps identify random coefficient parameters. Markets where
product characteristics are more spread out provide different "experiments" for
estimating how consumers substitute between products.

```{r rc_variation, eval=TRUE}
# Show price variation across markets for a few products
some_products <- c("F1B04", "F2B05", "F3B14")
for (p in some_products) {
  prices_p <- products$prices[products$product_ids == p]
  cat(sprintf("  %s: mean price = %.4f, sd = %.4f, range = [%.4f, %.4f]\n",
              p, mean(prices_p), sd(prices_p), min(prices_p), max(prices_p)))
}
```

### Setting Up Demographics

The Mixtape dataset includes demographic data with quarterly income for 20
simulated individuals per market. These demographics create observed
heterogeneity in price sensitivity: wealthier consumers are less price sensitive.

```{r rc_demographics, eval=TRUE}
demographics <- load_mixtape_demographics()

cat(sprintf("Demographic observations: %d\n", nrow(demographics)))
cat(sprintf("Markets: %d\n", length(unique(demographics$market))))
cat(sprintf("Individuals per market: %d\n",
            nrow(demographics) / length(unique(demographics$market))))

cat("\nQuarterly income summary:\n")
print(round(summary(demographics$quarterly_income), 1))

cat(sprintf("\nLog income range: [%.2f, %.2f]\n",
            min(log(demographics$quarterly_income)),
            max(log(demographics$quarterly_income))))
```

### Preparing Agent Data for rblp

To use demographics in rblp, we need to construct an agent data frame with
`market_ids`, `weights`, integration `nodes`, and demographic columns.

```{r rc_agent_setup, eval=FALSE}
# Prepare agent data with equal weights and log income
agents <- data.frame(
  market_ids = demographics$market,
  weights = 1 / 20,  # equal weights for 20 agents per market
  nodes0 = 0,        # placeholder node (not used for pure demographic model)
  log_income = log(demographics$quarterly_income),
  stringsAsFactors = FALSE
)

cat("Agent data preview:\n")
head(agents)
```

### Mushy x Log Income Interaction

Before estimating full random coefficients, the Mixtape exercises ask for a
simple demographic interaction: does the valuation of mushy cereals vary with
income? We add `log_income` as a demographic and interact it with `mushy` through
the pi matrix.

The pyblp solution finds that the pi coefficient on `mushy x log_income` is
approximately 0.251, indicating that higher-income consumers have a (slightly)
higher taste for mushy cereals.

```{r rc_mushy_income, eval=FALSE}
# Formulations
f1_rc <- blp_formulation(~ 0 + prices + mushy, absorb = ~ market_ids + product_ids)
f2_rc <- blp_formulation(~ 0 + mushy)

# Demographics formulation: log_income
f_demo <- blp_formulation(~ 0 + log_income)

# Create problem
rc_problem <- blp_problem(
  product_formulations = list(f1_rc, f2_rc),
  product_data = products,
  agent_formulation = f_demo,
  agent_data = agents
)

# Solve with pi on mushy x log_income
# sigma = 0 (no unobserved heterogeneity), pi = initial guess
pi0 <- matrix(0.1, nrow = 1, ncol = 1)  # K2=1 (mushy), D=1 (log_income)

rc_mushy_results <- rc_problem$solve(
  sigma = matrix(0, 1, 1),  # no sigma (fixed at 0)
  pi = pi0,
  method = "1s",
  optimization = blp_optimization("l-bfgs-b",
    method_options = list(maxit = 500))
)

print(rc_mushy_results)
```

**Expected result:** The pi coefficient on `mushy x log_income` is approximately
0.251, meaning a one-unit increase in log income raises the mean utility of
mushy cereals by about 0.25.

### Random Coefficient on Price (Sigma)

The next step adds a random coefficient (sigma) on price, allowing unobserved
heterogeneity in price sensitivity across consumers. Combined with the pi
interaction (price x income), this creates a rich model where price sensitivity
varies both with observed income and with unobserved consumer characteristics.

```{r rc_sigma_price, eval=FALSE}
# X2 now includes price (for sigma) and mushy (for pi)
f1_full <- blp_formulation(~ 0 + prices + mushy, absorb = ~ market_ids + product_ids)
f2_full <- blp_formulation(~ 0 + prices + mushy)

# Demographics: log_income
f_demo_full <- blp_formulation(~ 0 + log_income)

# Prepare products with proper IV
products_rc <- products  # demand_instruments0 already = price_instrument

rc_full_problem <- blp_problem(
  product_formulations = list(f1_full, f2_full),
  product_data = products_rc,
  agent_formulation = f_demo_full,
  agent_data = agents
)

# Starting values near the pyblp solution
# sigma: K2 x K2 = 2x2 diagonal (sigma on prices, 0 on mushy)
sigma0 <- diag(c(5.0, 0))  # start sigma_price at 5

# pi: K2 x D = 2x1 (prices x log_income, mushy x log_income)
pi0_full <- matrix(c(-5.0, 0.2), nrow = 2, ncol = 1)

rc_full_results <- rc_full_problem$solve(
  sigma = sigma0,
  pi = pi0_full,
  method = "1s",
  optimization = blp_optimization("l-bfgs-b",
    method_options = list(maxit = 1000, factr = 1e7))
)

print(rc_full_results)
```

**Expected results from pyblp:**

| Parameter | Estimate | Description |
|---|---|---|
| $\alpha$ (mean price) | negative | Mean price coefficient (in beta, from FE) |
| $\sigma_{\text{price}}$ | ~6.02 | SD of unobserved price sensitivity |
| $\pi_{\text{price} \times \text{income}}$ | ~-5.96 | Higher income $\Rightarrow$ less price sensitive |
| $\pi_{\text{mushy} \times \text{income}}$ | ~0.2 | Higher income $\Rightarrow$ slightly prefer mushy |

The large $\sigma_{\text{price}}$ (~6) indicates substantial unobserved
heterogeneity in price sensitivity, beyond what income explains. The negative
$\pi_{\text{price} \times \text{income}}$ means wealthier consumers are less
price sensitive, as expected.

### Why Random Coefficients Matter

With random coefficients, the substitution matrix is no longer governed by IIA.
Two products with similar characteristics (e.g., two mushy cereals at similar
prices) will have higher cross-price elasticities than two dissimilar products.
This has direct implications for:

- **Merger analysis:** Which mergers raise prices depends on which products are
  close substitutes
- **Counterfactual pricing:** Firms internalize cross-effects among their own products
- **Welfare analysis:** Consumer surplus calculations account for heterogeneous valuations

```{r rc_comparison, eval=FALSE}
# Compare logit vs RC logit elasticities
E_rc <- rc_full_results$compute_elasticities(first_market)

cat("=== Own-Price Elasticity Comparison (C01Q1) ===\n\n")
comp_table <- data.frame(
  product = mkt_data$product_ids[1:6],
  logit = sprintf("%.2f", diag(E)[1:6]),
  rc_logit = sprintf("%.2f", diag(E_rc)[1:6]),
  stringsAsFactors = FALSE
)
print(comp_table, row.names = FALSE, right = FALSE)
```

---

## Exercise 3: Micro Moments (Conceptual Overview)

Micro moments supplement the traditional BLP aggregate moments (based on market
shares) with information from individual-level data sources such as household
scanner panels, surveys, or browsing data.

### Motivation

In the standard BLP framework, identification of random coefficients comes from
cross-market variation in product sets, prices, and aggregate shares. This can
be weak when:

- Markets are similar in their product offerings
- The number of random coefficients exceeds what aggregate variation can identify
- We want to match specific demographic-choice correlations observed in micro data

Micro moments address this by directly targeting individual-level statistics.

### The Micro Moment Framework

A micro moment matches a statistic computed from micro data to its
model-predicted counterpart:

$$E\left[\bar{m}^{\text{micro}} - m(\theta)\right] = 0$$

where $\bar{m}^{\text{micro}}$ is the observed micro statistic (e.g., the
average income of consumers who buy mushy cereals) and $m(\theta)$ is the
model-predicted value (computed from the estimated choice probabilities and
demographic distributions).

### Types of Micro Moments

Common micro moments include:

1. **Conditional choice probabilities:** $E[y_{ij} | D_i \in G]$ -- the
   probability of buying product $j$ given demographics in group $G$
2. **Conditional means of characteristics:** $E[x_{j(i)} | D_i \in G]$ -- the
   mean characteristic of the chosen product for a demographic group
3. **Covariance moments:** $\text{Cov}(D_i, x_{j(i)})$ -- covariance between
   demographics and chosen product attributes

### Implementation in rblp

rblp provides three classes for constructing micro moments:

```{r micro_example, eval=FALSE}
# 1. Define a micro dataset (describes the data source)
md <- micro_dataset(
  name = "household scanner panel",
  observations = 50000,
  compute_weights = function(market_id, products, agents) {
    # Return I x J weight matrix
    I <- length(agents$weights)
    J <- length(products$shares)
    matrix(1, I, J)
  }
)

# 2. Define a micro part (what statistic to compute for each i,j pair)
mp <- micro_part(
  name = "income_x_mushy",
  dataset = md,
  compute_values = function(market_id, products, agents) {
    # Return I x J value matrix: income_i * mushy_j
    outer(agents$log_income, products$mushy)
  }
)

# 3. Define a micro moment (match target to prediction)
mm <- micro_moment(
  name = "E[log_income * mushy | purchase]",
  value = 0.5,   # observed from micro data
  parts = mp
)

# Pass to solve()
results <- rc_full_problem$solve(
  sigma = sigma0,
  pi = pi0_full,
  micro_moments = list(mm),
  method = "1s"
)
```

Micro moments can dramatically improve the precision of demographic interaction
parameters (pi) and random coefficient standard deviations (sigma) by providing
direct information about who buys what. For a full treatment, see Conlon and
Gortmaker (2020, Section 4) and the pyblp micro moments tutorial.

---

## Summary of Results

The table below summarizes key findings across all three exercises:

| Exercise | Model | Price Coeff. | Key Insight |
|---|---|---|---|
| 1 (Q3) | OLS logit | $-7.48$ | Biased toward zero by price endogeneity |
| 1 (Q4) | rblp OLS | $-7.48$ | Confirms equivalence of GMM/OLS |
| 1 (Q5) | FE logit (prices as IV) | $-28.6$ | FE absorb product quality; reveals true sensitivity |
| 1 (Q6) | IV + FE logit | $-30.6$ (SE $\approx$ 1.0) | IV corrects remaining endogeneity |
| 1 (Q7) | Counterfactual | -- | F1B04 share +230% from \$0.04 cut; IIA substitution |
| 1 (Q8) | Elasticities | -- | Own-price elasticities range $-2$ to $-5$ |
| 2 | Mushy x income | pi $\approx$ 0.25 | Wealthier consumers slightly prefer mushy |
| 2 | RC on price | $\sigma \approx 6.0$, $\pi \approx -6.0$ | Large heterogeneity in price sensitivity |
| 3 | Micro moments | -- | Individual data sharpens identification |

### Key Takeaways

1. **Price endogeneity is severe.** The OLS price coefficient ($-7.5$) is only
   one-quarter of the IV+FE estimate ($-30.6$). Failing to instrument biases the
   price effect toward zero, which leads to systematically understating
   consumer price sensitivity.

2. **Fixed effects matter.** Moving from OLS to FE logit (without IV) already
   moves the coefficient from $-7.5$ to $-28.6$. Most of the endogeneity bias
   in this application comes from product-level unobserved quality.

3. **Random coefficients break IIA.** The logit's substitution patterns are
   mechanical (proportional to shares). The mixed logit allows substitution to
   depend on product similarity, which matters for merger analysis, targeted
   pricing, and welfare evaluation.

4. **Demographics provide economic structure.** The interaction between price
   sensitivity and income ($\pi_{\text{price} \times \text{income}} \approx -6$)
   is both statistically and economically significant: it means that a
   low-income consumer faces four times the utility loss from a price increase
   compared to a high-income consumer.

---

## References

- Berry, S. (1994). Estimating Discrete-Choice Models of Product
  Differentiation. *RAND Journal of Economics*, 25(2), 242-262.

- Berry, S., Levinsohn, J., & Pakes, A. (1995). Automobile Prices in Market
  Equilibrium. *Econometrica*, 63(4), 841-890.

- Conlon, C. & Gortmaker, J. (2020). Best Practices for Differentiated Products
  Demand Estimation with pyblp. *RAND Journal of Economics*, 51(4), 1108-1161.

- Nevo, A. (2000). A Practitioner's Guide to Estimation of Random-Coefficients
  Logit Models of Demand. *Journal of Economics & Management Strategy*, 9(4),
  513-548.

- Mixtape Sessions. Demand Estimation.
  [https://github.com/Mixtape-Sessions/Demand-Estimation](https://github.com/Mixtape-Sessions/Demand-Estimation)
