---
title: "Random Coefficients Logit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random Coefficients Logit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rblp.verbose = FALSE)
```

## Introduction

The random coefficients (RC) logit -- also known as the mixed logit or BLP model -- is the workhorse of modern demand estimation for differentiated products. Unlike the plain logit, it generates flexible substitution patterns that are not subject to the IIA property.

This vignette covers:

1. The model and its parameters ($\sigma$, $\pi$)
2. Integration methods for simulating the mixing distribution
3. Estimation with and without demographics
4. GMM methods and convergence diagnostics

For the plain logit and nested logit, see `vignette("logit-nested-logit")`. For post-estimation analysis, see `vignette("post-estimation-mergers")`.

## The Random Coefficients Model

### Utility Specification

Consumer $i$ in market $t$ receives utility from product $j$:

$$u_{ijt} = \delta_{jt} + \mu_{ijt} + \varepsilon_{ijt}$$

where:

- $\delta_{jt} = x_{jt}'\beta + \alpha p_{jt} + \xi_{jt}$ is the **mean utility** (common to all consumers)
- $\mu_{ijt} = \sum_k x_{jkt}(\sigma_k \nu_{ik} + \pi_{kd} D_{id})$ is the **individual deviation** from the mean
- $\varepsilon_{ijt}$ is a Type I EV idiosyncratic shock

The individual-specific component $\mu_{ijt}$ has two sources of heterogeneity:

- **Unobserved heterogeneity** ($\sigma$): $\nu_{ik} \sim N(0, 1)$ are standard normal draws, and $\sigma_k$ captures the dispersion of tastes for characteristic $k$
- **Observed heterogeneity** ($\pi$): $D_{id}$ are observed demographics (income, age, etc.), and $\pi_{kd}$ captures how demographic $d$ shifts preferences for characteristic $k$

### The Sigma Matrix

The $\sigma$ parameter is a $K_2 \times K_2$ lower-triangular matrix (Cholesky root). In many applications, $\sigma$ is restricted to be diagonal, meaning that unobserved taste shocks for different characteristics are independent.

**Convention**: when specifying starting values for sigma:

- **Zero** on the diagonal means that parameter is **fixed at zero** (not estimated)
- **Non-zero** values are **starting values** for optimization

For example, with $K_2 = 4$ characteristics (intercept, prices, sugar, mushy):

```{r sigma_example, eval=FALSE}
# Diagonal sigma: all four RC variances are free
sigma <- diag(c(0.5, 3.0, 0.01, 0.1))

# Fix sugar and mushy RC at zero, estimate only intercept and price RC
sigma <- diag(c(0.5, 3.0, 0, 0))
```

### The Pi Matrix

When demographics are available, the $\pi$ matrix is $K_2 \times D$, where $D$ is the number of demographic variables. Element $\pi_{kd}$ captures how demographic $d$ shifts the taste coefficient for characteristic $k$.

## Integration Methods

The RC logit requires numerical integration over the distribution of unobserved heterogeneity $\nu$. The choice of integration method affects both accuracy and computation time.

### Product Rule (Gauss-Hermite)

The product rule uses Gauss-Hermite quadrature nodes, taking the tensor product across dimensions. With `size` nodes per dimension and $K_2$ random coefficients, this yields $\text{size}^{K_2}$ integration points per market.

```{r integration_product, eval=FALSE}
# 5 nodes per dimension, K2=4 -> 5^4 = 625 points per market
int_product <- blp_integration("product", size = 5)

# 3 nodes per dimension (faster, less accurate)
int_small <- blp_integration("product", size = 3)  # 3^4 = 81 points
```

The product rule is **deterministic** (no simulation noise) and provides exact integration for polynomials up to degree $2 \times \text{size} - 1$. However, it suffers from the curse of dimensionality: the number of points grows exponentially with $K_2$.

### Monte Carlo

Standard Monte Carlo draws random points from the mixing distribution. With `size` draws:

```{r integration_mc, eval=FALSE}
# 1000 Monte Carlo draws per market
int_mc <- blp_integration("monte_carlo", size = 1000, seed = 42)
```

Monte Carlo integration is flexible but introduces simulation noise that can cause roughness in the GMM objective function, making optimization difficult.

### Halton Sequences

Halton sequences are quasi-random low-discrepancy sequences that provide more uniform coverage than pseudo-random draws:

```{r integration_halton, eval=FALSE}
# 100 Halton draws per market (requires randtoolbox package)
int_halton <- blp_integration("halton", size = 100, seed = 42)
```

### Recommendation

For most applications:

- **Small $K_2$ (2--4)**: product rule with `size = 5` or `size = 7`
- **Moderate $K_2$ (5--8)**: Halton sequences with 100--500 points
- **Large $K_2$ (9+)**: Monte Carlo with 1000+ draws

## RC Logit without Demographics

The simplest RC logit has only unobserved heterogeneity (no demographics). This requires two formulations: one for the linear component ($X_1$) and one for the nonlinear component ($X_2$).

```{r rc_no_demo, eval=FALSE}
library(rblp)
products <- load_nevo_products()

# X1: linear demand (intercept + prices + sugar + mushy)
f1 <- blp_formulation(~ prices + sugar + mushy)

# X2: characteristics with random coefficients (same as X1 here)
f2 <- blp_formulation(~ prices + sugar + mushy)

# Create the problem with Gauss-Hermite integration
rc_problem <- blp_problem(
  product_formulations = list(f1, f2),
  product_data = products,
  integration = blp_integration("product", size = 3)  # 3^4 = 81 points
)

# Sigma starting values (4x4 diagonal)
initial_sigma <- diag(c(0.5, 0.5, 0.5, 0.5))

# Solve with one-step GMM
rc_results <- rc_problem$solve(
  sigma = initial_sigma,
  method = "1s",
  optimization = blp_optimization("l-bfgs-b",
    method_options = list(maxit = 200, factr = 1e7))
)

print(rc_results)
```

**Interpretation**:

- The estimated $\hat{\sigma}_k$ values capture the standard deviation of taste heterogeneity for each characteristic
- Large $\hat{\sigma}_{\text{price}}$ means consumers differ substantially in price sensitivity
- These taste differences generate realistic substitution: consumers who are less price-sensitive substitute toward premium products

## RC Logit with Demographics (Full Nevo Specification)

The full Nevo (2000) specification includes observed demographics (income, age, child presence) interacting with product characteristics. This is the canonical BLP specification.

```{r rc_demo, eval=FALSE}
library(rblp)
products <- load_nevo_products()
agents <- load_nevo_agents()

# Inspect the agent data
head(agents[, c("market_ids", "weights", "income", "income_squared", "age", "child")])

# X1: linear demand with product fixed effects (absorb product dummies)
# Only price is estimated; intercept, sugar, mushy are absorbed
f1 <- blp_formulation(~ 0 + prices, absorb = ~ product_ids)

# X2: nonlinear demand (intercept + prices + sugar + mushy get RC)
f2 <- blp_formulation(~ prices + sugar + mushy)

# Agent formulation: demographics that interact with X2
f_demo <- blp_formulation(~ 0 + income + income_squared + age + child)

# Create the problem (agent data provides nodes, weights, and demographics)
demo_problem <- blp_problem(
  product_formulations = list(f1, f2),
  product_data = products,
  agent_formulation = f_demo,
  agent_data = agents
)
```

The sigma and pi matrices encode the parameter structure:

```{r rc_demo_params, eval=FALSE}
# Sigma (4x4 diagonal): RC standard deviations
# Starting values near pyblp's converged estimates
sigma0 <- diag(c(0.558, 3.312, -0.006, 0.093))

# Pi (4x4): interactions between X2 characteristics and demographics
# Rows = X2 vars (intercept, prices, sugar, mushy)
# Cols = demographics (income, income_squared, age, child)
pi0 <- matrix(c(
  2.292, 0, 1.284, 0,        # intercept interactions
  588.3, -30.19, 0, 11.05,   # price interactions
  -0.384, 0, 0.0524, 0,      # sugar interactions
  0.748, 0, -1.354, 0        # mushy interactions
), nrow = 4, ncol = 4, byrow = TRUE)

# Solve
demo_results <- demo_problem$solve(
  sigma = sigma0,
  pi = pi0,
  method = "1s",
  optimization = blp_optimization("l-bfgs-b",
    method_options = list(maxit = 1000, factr = 1e7))
)

print(demo_results)
# Price coefficient ~ -63, matching pyblp's Nevo tutorial
```

**Key points about the Nevo specification**:

- The agent data includes 20 agents per market with pre-computed nodes (columns `nodes0`--`nodes3`), integration weights, and demographic variables
- Product fixed effects in $X_1$ absorb brand-level unobserved quality, so only the price coefficient is estimated in the linear part
- The large number of moment conditions (20 excluded instruments + exogenous characteristics) helps identify the many nonlinear parameters
- The Pi matrix zeros encode economic restrictions (e.g., `income_squared` only interacts with the price coefficient)

## Two-Step vs. One-Step GMM

### One-Step GMM (`method = "1s"`)

Uses the initial weighting matrix $W = (Z'Z/N)^{-1}$ (2SLS weight) throughout. This produces **consistent** estimates but is not asymptotically efficient.

### Two-Step GMM (`method = "2s"`)

1. **Step 1**: Estimate with the 2SLS weight $W_1 = (Z'Z/N)^{-1}$
2. **Update**: Compute the optimal weight $W^* = \hat{S}^{-1}$ from Step 1 residuals
3. **Step 2**: Re-estimate with $W^*$

Two-step GMM is **efficient** (achieves the smallest asymptotic variance among GMM estimators) but can exhibit finite-sample bias, especially with many moment conditions.

```{r gmm_methods, eval=FALSE}
# One-step (faster, consistent)
results_1s <- problem$solve(sigma = sigma0, method = "1s")

# Two-step (efficient, slower)
results_2s <- problem$solve(sigma = sigma0, method = "2s")
```

## Convergence Diagnostics

The RC logit objective function can be non-convex, so checking convergence is important.

### Optimizer Convergence

```{r diagnostics, eval=FALSE}
# Check if the optimizer converged
results$optimization_converged

# Number of iterations and function evaluations
results$optimization_iterations
results$optimization_evaluations

# Final objective value
results$objective

# Gradient at solution (should be near zero)
max(abs(results$gradient))
```

### Fixed-Point Convergence

The BLP contraction mapping must converge in every market at every evaluation:

```{r fp_diagnostics, eval=FALSE}
# Did all fixed points converge?
results$fp_converged

# Total contraction iterations (across all markets and evaluations)
results$fp_iterations
```

### Best Practices

1. **Try multiple starting values**: the GMM objective may have local minima
2. **Start from the logit solution**: use logit estimates as a sanity check
3. **Use tight contraction tolerances**: `blp_iteration("squarem")` is the default and is faster than the simple contraction
4. **Increase integration accuracy**: more nodes reduce simulation error
5. **Check the gradient norm**: it should be small at convergence

## Optimization Options

rblp supports several optimizers:

```{r optimizers, eval=FALSE}
# L-BFGS-B (default): quasi-Newton with bounds, uses gradient
opt_lbfgsb <- blp_optimization("l-bfgs-b",
  method_options = list(maxit = 1000, factr = 1e7))

# Nelder-Mead: derivative-free simplex method
opt_nm <- blp_optimization("nelder-mead",
  method_options = list(maxit = 5000))

# nlminb: PORT optimization, good with bounds
opt_nlminb <- blp_optimization("nlminb",
  method_options = list(iter.max = 1000, eval.max = 2000))
```

For most applications, **L-BFGS-B** is recommended because it uses analytical gradients (computed via the implicit function theorem) and respects parameter bounds.

## Iteration Options

The BLP contraction mapping can be accelerated with SQUAREM:

```{r iteration, eval=FALSE}
# SQUAREM (default): accelerated fixed-point iteration
iter_squarem <- blp_iteration("squarem")

# Simple contraction: safer but slower
iter_simple <- blp_iteration("simple",
  list(atol = 1e-14, max_evaluations = 5000))
```

SQUAREM typically converges in 10--50% fewer iterations than the simple contraction, with no loss of accuracy.

## Summary of Key Formulas

| Component | Formula |
|---|---|
| Mean utility | $\delta_{jt} = x_{jt}'\beta + \alpha p_{jt} + \xi_{jt}$ |
| Individual deviation | $\mu_{ijt} = \sum_k x_{jkt}(\sigma_k \nu_{ik} + \pi_{kd} D_{id})$ |
| Choice probability | $s_{ijt} = \frac{\exp(\delta_{jt} + \mu_{ijt})}{1 + \sum_k \exp(\delta_{kt} + \mu_{ikt})}$ |
| Predicted share | $s_{jt}(\delta, \theta) = \int s_{ijt} \, dF(\nu, D)$ |
| Contraction | $\delta^{r+1} = \delta^r + \log s^{\text{obs}} - \log s(\delta^r, \theta)$ |
| Moment condition | $E[Z_{jt}' \xi_{jt}(\theta)] = 0$ |
| GMM objective | $Q(\theta) = g(\theta)' W g(\theta)$, where $g = Z'\xi / N$ |

## References

- Berry, S., Levinsohn, J., & Pakes, A. (1995). Automobile Prices in Market Equilibrium. *Econometrica*, 63(4), 841-890.
- Nevo, A. (2000). A Practitioner's Guide to Estimation of Random-Coefficients Logit Models of Demand. *Journal of Economics & Management Strategy*, 9(4), 513-548.
- Conlon, C. & Gortmaker, J. (2020). Best Practices for Differentiated Products Demand Estimation with pyblp. *RAND Journal of Economics*, 51(4), 1108-1161.
- Varadhan, R. & Roland, C. (2008). Simple and Globally Convergent Methods for Accelerating the Convergence of Any EM Algorithm. *Scandinavian Journal of Statistics*, 35(2), 335-353.
