---
title: "Getting Started with rblp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rblp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rblp.verbose = FALSE)
```

## Overview

**rblp** is an R implementation of the BLP (Berry, Levinsohn, Pakes 1995) framework for estimating demand models for differentiated products. It is translated from the [pyblp](https://pyblp.readthedocs.io/) Python package by Conlon and Gortmaker (2020).

The package supports:

- **Plain logit** and **nested logit** demand estimation
- **Random coefficients logit** (mixed logit) with agent-level heterogeneity
- **Demographics** via Pi matrix interactions
- **Product fixed effects** via the `absorb` parameter
- **GMM estimation** (1-step and 2-step)
- **Post-estimation**: elasticities, consumer surplus, HHI, diversion ratios
- **Simulation**: generate equilibrium data and test estimation
- **Merger simulation**: compute post-merger prices and welfare effects

## Quick Start

```{r quickstart}
library(rblp)

# Load the Nevo (2000) cereal data
products <- load_nevo_products()
head(products[, c("market_ids", "product_ids", "shares", "prices", "sugar", "mushy")])
```

The data has `r nrow(products)` product-market observations across `r length(unique(products$market_ids))` markets. The `demand_instruments0` through `demand_instruments19` columns contain 20 pre-computed BLP-style excluded instruments.

## Example 1: Plain Logit

The simplest demand model. We regress the log odds ratio $\delta_{jt} = \log(s_{jt}) - \log(s_{0t})$ on product characteristics using IV-GMM.

```{r logit}
# Define the linear model: intercept + prices + sugar + mushy
f1 <- blp_formulation(~ prices + sugar + mushy)

# Create and solve the problem
logit_problem <- blp_problem(list(f1), products)
logit_results <- logit_problem$solve(method = "1s")

print(logit_results)
```

The price coefficient is negative as expected -- higher prices reduce utility. The exogenous characteristics (sugar, mushy) serve as instruments for prices.

## Example 2: Logit with Product Fixed Effects

This matches the pyblp Nevo tutorial specification. Product fixed effects absorb all time-invariant product characteristics (brand, sugar, mushy, etc.), so we only estimate the price coefficient.

```{r fe_logit}
# Absorb product-level fixed effects (Frisch-Waugh-Lovell demeaning)
f1_fe <- blp_formulation(~ prices, absorb = ~ product_ids)

fe_problem <- blp_problem(list(f1_fe), products)
fe_results <- fe_problem$solve(method = "1s")

print(fe_results)
```

The price coefficient of `r round(fe_results$beta[1], 2)` matches pyblp's reported value of approximately -30.

## Example 3: Random Coefficients Logit

The random coefficients (RC) logit allows consumer heterogeneity in preferences. Each consumer $i$ has individual-specific taste coefficients:

$$u_{ijt} = \delta_{jt} + \sum_k x_{jk} \sigma_k \nu_{ik} + \varepsilon_{ijt}$$

where $\nu_{ik}$ are standard normal draws and $\sigma_k$ captures the dispersion of tastes.

```{r rc_logit, eval=FALSE}
# X1: linear demand characteristics
f1_rc <- blp_formulation(~ prices + sugar + mushy)
# X2: random coefficient characteristics (same as X1 here)
f2_rc <- blp_formulation(~ prices + sugar + mushy)

rc_problem <- blp_problem(
  product_formulations = list(f1_rc, f2_rc),
  product_data = products,
  integration = blp_integration("product", size = 3)  # Gauss-Hermite quadrature
)

# Sigma: diagonal matrix with starting values (0 = fixed, non-zero = free)
initial_sigma <- diag(c(0.5, 0.5, 0.5, 0.5))

rc_results <- rc_problem$solve(
  sigma = initial_sigma,
  method = "1s",
  optimization = blp_optimization("l-bfgs-b",
    method_options = list(maxit = 200, factr = 1e7))
)

print(rc_results)
```

**Key parameters:**

- `integration = blp_integration("product", size = 3)` uses Gauss-Hermite product rule with 3 nodes per dimension ($3^4 = 81$ integration points per market)
- `sigma` is the K2 x K2 Cholesky root. Zeros on the diagonal fix that parameter; non-zero values are starting values for optimization
- `method = "2s"` runs two-step efficient GMM (default); `"1s"` runs one-step

## Example 4: RC Logit with Demographics

Demographics allow observed consumer characteristics to interact with product characteristics via the Pi matrix:

$$u_{ijt} = \delta_{jt} + \sum_k x_{jk} (\sigma_k \nu_{ik} + \sum_d \pi_{kd} D_{id}) + \varepsilon_{ijt}$$

```{r rc_demo, eval=FALSE}
# Load agent (consumer) data with demographics
agents <- load_nevo_agents()
head(agents[, c("market_ids", "weights", "income", "age", "child")])

# Demographics formulation (no intercept)
demo_form <- blp_formulation(~ 0 + income + income_squared + age + child)

demo_problem <- blp_problem(
  product_formulations = list(
    blp_formulation(~ prices + sugar + mushy),  # X1
    blp_formulation(~ prices + sugar + mushy)   # X2
  ),
  product_data = products,
  agent_formulation = demo_form,
  agent_data = agents
)

# Sigma (4x4 diagonal) and Pi (4x4, K2 x D)
# Starting values from Nevo's published estimates
sigma0 <- diag(c(0.3302, 2.4526, 0.0163, 0.2441))
pi0 <- matrix(c(
  5.4819, 0, 0.2037, 0,
  15.8935, -1.2000, 0, 2.6342,
  -0.2506, 0, 0.0511, 0,
  1.2650, 0, -0.8091, 0
), nrow = 4, ncol = 4, byrow = TRUE)

demo_results <- demo_problem$solve(
  sigma = sigma0, pi = pi0,
  method = "1s",
  optimization = blp_optimization("l-bfgs-b",
    method_options = list(maxit = 500, factr = 1e7))
)

print(demo_results)
```

## Post-Estimation

After estimation, compute policy-relevant quantities:

### Elasticities

```{r elasticities}
# Compute elasticities for the first market
first_market <- logit_results$problem$unique_market_ids[1]
E <- logit_results$compute_elasticities(first_market)

# Own-price elasticities (diagonal)
cat("Own-price elasticities (first 5 products):\n")
print(round(diag(E)[1:5], 3))
```

### Consumer Surplus

```{r cs}
cs <- logit_results$compute_consumer_surplus()
cat("Consumer surplus (first 5 markets):\n")
print(round(cs[1:5], 4))
```

### HHI

```{r hhi}
hhi <- logit_results$compute_hhi()
cat("HHI (first 5 markets):\n")
print(round(hhi[1:5], 0))
```

## Simulation

Generate synthetic equilibrium data to test the estimator:

```{r simulation}
# Create a balanced panel: 50 markets, 20 products/market, 4 firms
id_data <- build_id_data(T = 50, J = 20, F = 4)
set.seed(42)
id_data$x <- runif(nrow(id_data), 0, 1)

# True parameters: beta = (intercept=0.5, price=-2, x=0.8)
sim <- blp_simulation(
  product_formulations = list(blp_formulation(~ prices + x)),
  product_data = id_data,
  beta = c(0.5, -2, 0.8),
  xi_variance = 0.3,
  seed = 42
)

# Solve for equilibrium prices and shares
sim_results <- sim$replace_endogenous(
  iteration = blp_iteration("simple", list(atol = 1e-12, max_evaluations = 5000))
)

cat(sprintf("Price range: [%.3f, %.3f]\n", min(sim_results$prices), max(sim_results$prices)))
cat(sprintf("Share range: [%.5f, %.5f]\n", min(sim_results$shares), max(sim_results$shares)))
```

Now estimate back the true parameters:

```{r sim_estimate}
sim_problem <- sim_results$to_problem()
sim_est <- sim_problem$solve(method = "1s")

cat("True vs Estimated:\n")
cat(sprintf("  intercept: true=0.50, est=%.4f\n", sim_est$beta[1]))
cat(sprintf("  price:     true=-2.00, est=%.4f\n", sim_est$beta[2]))
cat(sprintf("  x:         true=0.80, est=%.4f\n", sim_est$beta[3]))
```

## Merger Simulation

Simulate the effect of a merger between firms 2 and 3:

```{r merger}
new_firm_ids <- sim_results$product_data$firm_ids
new_firm_ids[new_firm_ids == 3] <- 2  # merge firm 3 into firm 2

merger <- sim_est$compute_merger(
  new_firm_ids = new_firm_ids,
  iteration = blp_iteration("simple", list(atol = 1e-12, max_evaluations = 5000))
)

cat(sprintf("Mean price change: %.2f%%\n", mean(merger$price_change_pct)))
cat(sprintf("Max price change:  %.2f%%\n", max(merger$price_change_pct)))
cat(sprintf("Mean CS change:    %.4f\n", mean(merger$delta_cs)))
```

## Key Classes and Functions

| Function | Purpose |
|---|---|
| `blp_formulation()` | Define model formulas (with optional `absorb` for FE) |
| `blp_problem()` | Create an estimation problem from data and formulations |
| `blp_simulation()` | Create a simulation with known parameters |
| `blp_integration()` | Configure integration (product rule, Monte Carlo, etc.) |
| `blp_iteration()` | Configure fixed-point iteration (SQUAREM, simple, etc.) |
| `blp_optimization()` | Configure optimization (L-BFGS-B, Nelder-Mead, etc.) |
| `build_blp_instruments()` | Construct BLP-style instruments |
| `build_differentiation_instruments()` | Gandhi-Houde differentiation instruments |
| `build_id_data()` | Create balanced panel identifiers |

## Comparison with pyblp

rblp aims to replicate pyblp's functionality in R. Key correspondence:

| pyblp | rblp |
|---|---|
| `Formulation('prices', absorb='C(product_ids)')` | `blp_formulation(~ prices, absorb = ~ product_ids)` |
| `Problem(formulation, product_data)` | `blp_problem(list(f1), product_data)` |
| `problem.solve()` | `problem$solve()` |
| `results.compute_elasticities()` | `results$compute_elasticities()` |
| `Integration('product', size=5)` | `blp_integration("product", size = 5)` |
| `Simulation(...)` | `blp_simulation(...)` |

## References

- Berry, S., Levinsohn, J., & Pakes, A. (1995). Automobile Prices in Market Equilibrium. *Econometrica*, 63(4), 841-890.
- Nevo, A. (2000). A Practitioner's Guide to Estimation of Random-Coefficients Logit Models of Demand. *Journal of Economics & Management Strategy*, 9(4), 513-548.
- Conlon, C. & Gortmaker, J. (2020). Best Practices for Differentiated Products Demand Estimation with pyblp. *RAND Journal of Economics*, 51(4), 1108-1161.
