---
title: "Simulation and Monte Carlo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation and Monte Carlo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rblp.verbose = FALSE)
```

## Introduction

Simulation is an essential tool for validating estimators and understanding their properties. The rblp package provides a complete simulation framework:

1. **Specify** a data-generating process (DGP) with known parameters
2. **Solve** for equilibrium prices and market shares
3. **Estimate** the model from the simulated data
4. **Compare** estimated parameters to the true values

This vignette covers single-run roundtrip validation and discusses Monte Carlo study design. For estimation details, see `vignette("logit-nested-logit")` and `vignette("random-coefficients")`.

## The BLPSimulation Class

The `blp_simulation()` function creates a simulation object that holds the true parameters and generates structural errors. The key method `replace_endogenous()` then solves for the Nash-Bertrand equilibrium prices and the resulting market shares.

### Workflow

```
build_id_data() -> add characteristics -> blp_simulation() -> replace_endogenous() -> to_problem() -> solve()
```

## Step 1: Build the Panel Structure

The `build_id_data()` function creates a balanced panel with specified numbers of markets, products per market, and firms:

```{r build_panel}
library(rblp)

# 50 markets, 20 products per market, 4 firms
id_data <- build_id_data(T = 50, J = 20, F = 4)

cat(sprintf("Total observations: %d\n", nrow(id_data)))
cat(sprintf("Markets: %d\n", length(unique(id_data$market_ids))))
cat(sprintf("Products per market: %d\n",
            nrow(id_data) / length(unique(id_data$market_ids))))

# Products per firm
table(id_data$firm_ids[id_data$market_ids == 1])
```

Products are distributed as evenly as possible across firms. If $J$ is not divisible by $F$, the first firms get one extra product.

### Adding Product Characteristics

Product characteristics must be added to the panel before simulation. These should be exogenous (independent of the structural error $\xi$):

```{r add_chars}
set.seed(42)

# Add an exogenous product characteristic
id_data$x <- runif(nrow(id_data), 0, 1)

# Can add more characteristics
id_data$z <- rnorm(nrow(id_data), 0, 0.5)

head(id_data)
```

## Step 2: Specify the DGP and Simulate

### Logit DGP

The simplest DGP is a logit model with known linear parameters:

```{r logit_sim}
# True parameters: beta = (intercept, price_coeff, x_coeff)
sim <- blp_simulation(
  product_formulations = list(blp_formulation(~ prices + x)),
  product_data = id_data,
  beta = c(0.5, -2, 0.8),    # true beta
  xi_variance = 0.3,          # variance of demand shocks
  seed = 42
)

print(sim)
```

### Solving for Equilibrium

The `replace_endogenous()` method computes equilibrium prices and shares by iterating the Bertrand pricing equation:

$$p = mc + \Omega(p)^{-1} s(p)$$

where $\Omega$ is the ownership-weighted matrix of share derivatives and $s(p)$ are the logit shares evaluated at prices $p$.

```{r equilibrium}
sim_data <- sim$replace_endogenous(
  iteration = blp_iteration("simple", list(atol = 1e-12, max_evaluations = 5000))
)

cat(sprintf("Equilibrium price range:  [%.4f, %.4f]\n",
            min(sim_data$prices), max(sim_data$prices)))
cat(sprintf("Equilibrium share range:  [%.6f, %.6f]\n",
            min(sim_data$shares), max(sim_data$shares)))
cat(sprintf("Mean price: %.4f\n", mean(sim_data$prices)))
cat(sprintf("Mean share: %.6f\n", mean(sim_data$shares)))
```

The `BLPSimulationResults` object contains the equilibrium product data (with updated prices and shares), the marginal costs, and the mean utilities.

## Step 3: Estimate from Simulated Data

The `to_problem()` method converts the simulation results into a `BLPProblem`, automatically constructing BLP instruments from the exogenous characteristics:

```{r estimate}
sim_problem <- sim_data$to_problem()
sim_results <- sim_problem$solve(method = "1s")

print(sim_results)
```

### Comparing to True Values

```{r compare}
true_beta <- c(0.5, -2, 0.8)
est_beta <- sim_results$beta

comparison <- data.frame(
  parameter = c("intercept", "price", "x"),
  true = true_beta,
  estimated = round(est_beta, 4),
  bias = round(est_beta - true_beta, 4),
  bias_pct = round((est_beta - true_beta) / abs(true_beta) * 100, 2)
)

print(comparison, row.names = FALSE)
```

The estimates should be close to the true values, confirming that the estimator is working correctly. Small deviations are expected due to finite-sample variation.

## Instrument Construction

The `to_problem()` method automatically constructs BLP instruments when the product data does not contain pre-computed instruments. You can also construct instruments manually.

### BLP Instruments

Traditional BLP instruments use sums of characteristics of own-firm and rival products:

```{r blp_instruments}
# Extract the exogenous characteristic
X_exog <- as.matrix(sim_data$product_data$x)

# Build BLP instruments
blp_iv <- build_blp_instruments(
  X = X_exog,
  market_ids = sim_data$product_data$market_ids,
  firm_ids = sim_data$product_data$firm_ids
)

cat(sprintf("BLP instruments: %d columns\n", ncol(blp_iv)))
cat("Column names:", paste(colnames(blp_iv), collapse = ", "), "\n")
```

For each characteristic, BLP instruments produce two columns:

- **own**: sum of the characteristic for other products of the same firm
- **rival**: sum of the characteristic for products of rival firms

These are valid instruments because they are correlated with prices (through the markup equation) but uncorrelated with unobserved quality $\xi$ (since characteristics are exogenous).

### Differentiation Instruments

Gandhi and Houde (2020) propose instruments based on product proximity in characteristic space:

```{r diff_instruments}
# Local differentiation instruments
diff_iv_local <- build_differentiation_instruments(
  X = X_exog,
  market_ids = sim_data$product_data$market_ids,
  firm_ids = sim_data$product_data$firm_ids,
  method = "local"
)

# Quadratic differentiation instruments
diff_iv_quad <- build_differentiation_instruments(
  X = X_exog,
  market_ids = sim_data$product_data$market_ids,
  firm_ids = sim_data$product_data$firm_ids,
  method = "quadratic"
)

cat(sprintf("Local diff. instruments: %d columns\n", ncol(diff_iv_local)))
cat(sprintf("Quadratic diff. instruments: %d columns\n", ncol(diff_iv_quad)))
```

The **local** method counts the number of rival products within one standard deviation of each characteristic, while the **quadratic** method sums the squared distances.

Differentiation instruments can improve identification of random coefficients because they capture the degree of product isolation, which directly affects markups through the substitution patterns.

## Random Coefficients Simulation

For RC logit simulation, specify the sigma matrix (and optionally pi and integration):

```{r rc_sim, eval=FALSE}
# RC logit DGP with 2 random coefficients
id_data2 <- build_id_data(T = 30, J = 15, F = 3)
set.seed(123)
id_data2$x <- runif(nrow(id_data2), 0, 1)

rc_sim <- blp_simulation(
  product_formulations = list(
    blp_formulation(~ prices + x),  # X1: linear
    blp_formulation(~ x)            # X2: nonlinear (RC on x only)
  ),
  product_data = id_data2,
  beta = c(1.0, -3.0, 1.5),   # true linear parameters
  sigma = matrix(0.8),         # true sigma for x (1x1)
  integration = blp_integration("product", size = 5),
  xi_variance = 0.5,
  seed = 123
)

# Solve for equilibrium
rc_sim_data <- rc_sim$replace_endogenous(
  iteration = blp_iteration("simple", list(atol = 1e-12, max_evaluations = 5000))
)

# Estimate
rc_sim_problem <- rc_sim_data$to_problem()
rc_sim_results <- rc_sim_problem$solve(
  sigma = matrix(0.5),  # starting value (not the true value)
  method = "1s",
  optimization = blp_optimization("l-bfgs-b",
    method_options = list(maxit = 200))
)

# Compare
cat("True vs. Estimated:\n")
cat(sprintf("  beta:  true = (1.0, -3.0, 1.5), est = (%s)\n",
            paste(round(rc_sim_results$beta, 3), collapse = ", ")))
cat(sprintf("  sigma: true = 0.8, est = %.3f\n", rc_sim_results$sigma[1, 1]))
```

## Monte Carlo Study Design

A Monte Carlo study repeats the simulate-estimate cycle many times to study the **finite-sample properties** of the estimator: bias, variance, coverage of confidence intervals, etc.

### Design Considerations

1. **Number of replications**: 500--1000 is typical for asymptotic properties; 100 may suffice for debugging
2. **DGP parameters**: choose values that produce realistic shares and prices
3. **Market/product structure**: vary $T$ (markets), $J$ (products/market), $F$ (firms) to study how sample size affects performance
4. **Instrument strength**: compare BLP vs. differentiation instruments
5. **Integration accuracy**: vary the number of nodes to study simulation error
6. **Seeds**: use a different seed for each replication but make results reproducible

### Template for a Monte Carlo Study

```{r monte_carlo, eval=FALSE}
# Monte Carlo study: logit estimator performance
set.seed(2024)
n_reps <- 100
true_beta <- c(0.5, -2, 0.8)

# Storage for results
estimates <- matrix(NA, n_reps, 3)
colnames(estimates) <- c("intercept", "price", "x")

for (rep in seq_len(n_reps)) {
  # 1. Generate data
  id_data_mc <- build_id_data(T = 50, J = 20, F = 4)
  id_data_mc$x <- runif(nrow(id_data_mc), 0, 1)

  sim_mc <- blp_simulation(
    product_formulations = list(blp_formulation(~ prices + x)),
    product_data = id_data_mc,
    beta = true_beta,
    xi_variance = 0.3,
    seed = rep  # different seed each replication
  )

  sim_data_mc <- sim_mc$replace_endogenous(
    iteration = blp_iteration("simple",
      list(atol = 1e-12, max_evaluations = 5000))
  )

  # 2. Estimate
  sim_prob_mc <- sim_data_mc$to_problem()
  sim_res_mc <- sim_prob_mc$solve(method = "1s")

  # 3. Store
  estimates[rep, ] <- sim_res_mc$beta
}

# 4. Summarize
cat("Monte Carlo Results (", n_reps, " replications)\n")
cat("========================================\n")
for (k in seq_len(3)) {
  cat(sprintf("  %s: true = %.2f, mean = %.4f, bias = %.4f, RMSE = %.4f\n",
              colnames(estimates)[k],
              true_beta[k],
              mean(estimates[, k]),
              mean(estimates[, k]) - true_beta[k],
              sqrt(mean((estimates[, k] - true_beta[k])^2))))
}
```

### What to Report

A Monte Carlo study should report:

| Metric | Formula | Interpretation |
|---|---|---|
| Mean bias | $\bar{\hat{\beta}} - \beta_0$ | Systematic deviation from truth |
| Median bias | $\text{median}(\hat{\beta}) - \beta_0$ | Robust to outliers |
| RMSE | $\sqrt{E[(\hat{\beta} - \beta_0)^2]}$ | Overall accuracy |
| Std. dev. | $\text{sd}(\hat{\beta})$ | Precision of estimator |
| Coverage | Fraction of CIs containing $\beta_0$ | Should be near 95% for 95% CIs |
| Rejection rate | Fraction of t-tests rejecting $H_0: \beta = \beta_0$ | Should be near 5% at 5% level |

## Supply-Side Simulation

The simulation framework also supports supply-side estimation with a cost function. This requires a third formulation (X3) and supply instruments:

```{r supply_sim, eval=FALSE}
# DGP with supply side
id_data_supply <- build_id_data(T = 30, J = 15, F = 3)
set.seed(99)
id_data_supply$x <- runif(nrow(id_data_supply), 0, 1)
id_data_supply$w <- runif(nrow(id_data_supply), 0.5, 1.5)  # cost shifter

supply_sim <- blp_simulation(
  product_formulations = list(
    blp_formulation(~ prices + x),   # X1: demand
    blp_formulation(~ x),            # X2: RC on x
    blp_formulation(~ x + w)         # X3: cost function
  ),
  product_data = id_data_supply,
  beta = c(1.0, -3.0, 1.5),
  sigma = matrix(0.5),
  gamma = c(0.5, 0.3, 0.8),   # true cost parameters
  integration = blp_integration("product", size = 5),
  xi_variance = 0.3,
  omega_variance = 0.2,
  correlation = 0.7,  # correlation between xi and omega
  seed = 99
)

supply_data <- supply_sim$replace_endogenous()

# The resulting problem will include supply-side estimation
supply_problem <- supply_data$to_problem()
```

## Summary

The simulation framework in rblp provides a complete toolkit for:

- **Validation**: confirm that the estimator recovers known parameters
- **Power analysis**: how many markets/products are needed for acceptable precision?
- **Instrument comparison**: which instruments perform better in finite samples?
- **Robustness**: how sensitive are results to DGP assumptions?

| Function | Purpose |
|---|---|
| `build_id_data(T, J, F)` | Create balanced panel structure |
| `blp_simulation(...)` | Specify DGP with true parameters |
| `sim$replace_endogenous()` | Solve for equilibrium prices/shares |
| `sim_data$to_problem()` | Convert to estimation problem |
| `build_blp_instruments()` | Traditional BLP instruments |
| `build_differentiation_instruments()` | Gandhi-Houde instruments |

## References

- Berry, S., Levinsohn, J., & Pakes, A. (1995). Automobile Prices in Market Equilibrium. *Econometrica*, 63(4), 841-890.
- Gandhi, A. & Houde, J.-F. (2020). Measuring Substitution Patterns in Differentiated-Products Industries. NBER Working Paper.
- Conlon, C. & Gortmaker, J. (2020). Best Practices for Differentiated Products Demand Estimation with pyblp. *RAND Journal of Economics*, 51(4), 1108-1161.
